Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2019-02-04 Response_referees.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

List of major changes:
\begin{itemize}
 \item We have further clarified the relation between localization properties and features of the computational basis needed
 to provide precise results. Major modifications under this perspective have been done in the introduction, in section II at pag. 5 before the analysis of fig. 2,in section IV and in the conclusions.
 \item The introduction has been deeply revised. In particular we have reduced the length of the first two paragraph concerning the differences between accuracy and precision.
 \item We added a comment after eq. (7) on the non-hermiticity of the response density and corrected the equation (11).
 \item Added a dedicated subsection (actually the IIA at page 4)  to discuss the localization properties of the FS states for static perturbations.
 \item Added a panel in figure 2 that contains the plot of the real part of $\alpha(\omega)$ in calculated in different computational domains. This new plot aims to evidence that the localization
 properties of th FS affect, on equal footing, both the real and the imaginary part of the linear response functional.
 \item We have reorganized the material after section III A. In particular we have introduced a new section (number IV in the actual version of the paper) in which we have discussed both the physical features of the excitation spectrum and the relative importance of localized vs. delocalized sector in expressing the susceptibility functional. This section also collects computational results for molecules beyond the case of $CO$ and benzene.
 \item We have also included in our analysis the treatment of various molecular systems, of various sizes and symmetries, such as to further illustrate the generality of our assumptions. A table has been added for comparing the excitation landscape of the molecules, in particular related to the ability of the numerical excitations to
 express the static polarizability tensor.
 \item we have added a reference to the URL of the supplementary material which contains the data analysis and further results in support of our findings
\end{itemize}

@@

  Nonetheless, to have an idea of the dimension of the box needed to achieve convergence we can refer to the case of the Benzene molecule and make usage
  of the result found in our analysis, compared to the one reported in fig. 2 of \cite{baroni2008}. Within our analysis we show that a box linear size of about 18 Angostroem is sufficient to produce converged results for excitation spectrum below IP (around 6.5 eV in the present case). On the contrary, results of \cite{baroni2008} evidence that to achieve convergence up to 30 eV a box of linear dimension of more than 30 Angstroem is needed, so going from the localized to the delocalized sector of excitation spectrum requires a relevant increase in the simulation domain.

  \item \emph{on the point concerning the convergence rate of the delocalized vs. localized states with respect
  to the grid spacing}

 In presenting the results of this paper care has been taken to ensure the convergence of the results wrt the computational parameters. We have found that there are no relevant difference in the convergence rate of localized vs delocalized quantities (for instance excitations) in function of the grid spacing. This is due to the fact that localized and delocalized states differ in their asymptotic properties whereas the grid spacing convergence is influenced by the short-range oscillation of the functions.

 The values of the grid spacing used for the calculations may be found in the supplementary material.

@@

\begin{enumerate}

 \item \emph{On the localized sector of excitation. Dependence on the symmetry and dimension of the molecule.}

 % In section IIIA we have provided formal arguments that evidence that the excitation spectrum splits into two sectors with distinct features. The first one
 % is composed by a finite set of localized excitations with discrete eigenvalues, whereas the second one contains a continuum spectrum of delocalized objects.
 % For excitation energies lower than the ionization potential (IP) of the molecule all the excitations belong to the first sector. In the intermediate energy range which goes from IP to the absolute value of the energy of the deepest occupied orbital the two sectors compenetrate so that there are localized excitation embedded in a continuum of delocalized ones, and finally only elements of the second sector exist for higher values of the energy.
 %
 % These formal results have been supported by numerical calculations in which the excitation spectrum of several molecules has been explicitly computed.
 We have added in the Appendix the explaination
 of the procedure we employed for the determination of the excitation landscape.
  In the previous version of the paper this type of analysis was limited to the case of CO and Benzene molecules.

  However the same structure of the excitation spectrum is quite general and turns out not to be affected by the size or symmetry of the molecule. To check this statement we have repeated
 the same computational analysis for several other molecules of different shape and dimension like the Fullerene and the Aflatoxin, a low-symmetry molecule of 35 atoms which is not as little and exhibit generic features. Results for this studies
 are presented as supplementary materials. See Ref.22 of the revised manuscript.

 Looking at the excitation landscape for these cases, extracted with the same procedure and tolerance parameters of the cases presented in the paper we see that our consideration are relatively independent from the molecule symmetry: in particular, the number of excitations which
 belong to the localized sector, even above the first IP
 seems to increase with the molecule's size.
 This is due to the increasing of the KS density of states below the continuum threshold which ensures the realization of more localized excitations.

### 2019-02-04 AnalyticStructure_main.tex

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response (LR) formalism in molecules,
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
We show that for frequencies below the first ionization potential (IP) of the system, it is possible to express the response density by employing localized states only.
Above this threshold energy, such a locality property cannot be achieved.
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations -- which may also extend above IP -- that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness.
We present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated
in any discretization scheme.
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models'
approximations on the scientific outcomes of the simulation.
\end{abstract}

@@

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictivity power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the found discrepancy.
For communities employing Density Functional Theory (DFT), the need of such a ``calibration'' of computer codes has led to the DeltaCodes project~\cite{deltaTest2016}.
This project represents a remarkable example of precision-driven initiative where a grassroot community of code-owners struggled to extract
well-defined computational results in the context of DFT calculations. Ground-State quantum mechanical quantities
as lattice constants and bulk modulus of elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that \emph{open} (i.e. Isolated) Boundary Conditions (BC), that are imposed to the system, will have on the computational evaluation of LR quantities.
We perform analytic considerations on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others that would \emph{explicitly} depend of the employed computational treatment.

@@

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed
for ground state calculations of molecules.
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in Appendix~\ref{compdetails}.

@@

Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that will turn out to be very useful
in the forthcoming considerations about linear-response quantities. In particular we revisit the impact that the Isolated BC have on the solutions
of the Schr\"odinger Equation.

@@

When describing the Ground State of a molecule, we suppose that the potential operator, even if non-local,
can be effectively restricted on wavefunctions that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite}
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though
perhaps in a larger region.
For a computational analysis of the problem, this property is fundamental in the choice of the numerical treatment as we know that
computational basis sets
that are tailored to express asymptotically vanishing wavefunctions are better suited for the solution of the problem.

@@

From the above expressions we can see that, for molecular systems,the value $\epsilon=0$
represents a \emph{threshold}~\footnote{The threshold value in general should correspond to asymptotic value of the potential at infinity.}
energy that separates two different classes of solutions.
Negative-energy solutions are \emph{localized}, i.e. they exhibit bound-state behavior far away from the scatterer.
These are bound states and, by means of their normalizability, they are associated to \emph{discrete}, well-defined energies.
The bound states can be of course labeled by quantum numbers that are defined by the properties of the scatterer, namely the
symmetries of the potential $\op V$.
In a computational discretization of the Schr\"odinger equation, such states can be in principle expressed with arbitrary precision,
provided that the numerical treatment offers the adequate level of completeness.
Being discrete and well-identified, the bound-state energies are a property of the molecule and can be associated to \emph{observable}
quantities. The values of their energies can be compared among different treatments and it is in principle possible to
provide reference results on them.

@@

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}, where the authors analyze the effects of a finite basis set on the
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character,
and the independence of its energy with respect to the basis set.
Furthermore, it is clearly shown that the energies of the unbound orbitals are not stable with respect to parameters like the size of the computational domain.

\section{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}
It is evident that the arguments presented above have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems that are electronically stable (i.e. no anionic or metastable configurations),
such quantity is localized, in the sense of the above section. Stated otherwise,
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
In this framework one considers the unperturbed Hamiltonian $\hnot$ and assumes that
the GS density $\dmnot$ is accessible and expressed in a given computational treatment.
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system.
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part $\Liouvnot$ plus a coupling term $ \coupl$. The action of these
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidences that the response density acts as a transition
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfy the \emph{transverse} condition
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in empty subspace of $\hnot$
(for more details see Appendix~\ref{LiouvillianAction}).
In other terms, the condition $\dm'_\Phi = \dm'_{\Phi\,\perp}$ holds.

@@

From equation \eqref{rhoPrimeFluctuationStateDef2} we see that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is spatially localized, \textit{i.e.} its evaluation can be restricted on a bounded real-space domain.
Even when such locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

@@

\subsection{Localized behavior for static perturbations}
The locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain.
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem.
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques,
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{Fig1_CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line)
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest
box.}
\end{figure}

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strengths $F_j$ small enough to
preserve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behavior of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
here we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

Two distinct regimes can be identified, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment, as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

@@

These considerations prove why, when a given computer code employs established GS numerical techniques,
\emph{it is much easier to converge LR quantities for $\omega$ below IP}.
Above the IP threshold, a computational setup
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able (even implicitly) to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behavior in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.
The IP threshold has an evident physical interpretation: for energies in a range higher than the ionization potential,
localized and delocalized states might be coupled by the perturbation. Obtaining convergence of the observable quantities at these energies is therefore much more challenging if one employs
numerical techniques which were originally tailored for GS calculations.

>>>>

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response (LR) formalism in molecules,
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
We show that for frequencies below the first ionization potential (IP) of the system, it is possible to express the response density by employing localized states only.
Above this threshold energy, such a locality property cannot be achieved.
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations -- which may also extend above IP -- that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness.
We present indicators that can help to quantify such potential observable properties of an excitation, that can be evaluated
in any discretization scheme.
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models'
approximations on the scientific outcomes of the simulation.
\end{abstract}

@@

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictive power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the observed discrepancy.
For communities employing Density Functional Theory (DFT), the need of such a ``calibration'' of computer codes has led to the DeltaCodes project~\cite{deltaTest2016}.
This project represents a remarkable example of precision-driven initiative where a grassroot community of code-owners struggled to extract
well-defined computational results in the context of DFT calculations. Ground-State quantum mechanical quantities
as lattice constants and bulk modulus of elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.

@@

In this paper, we focus our attention on Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that \emph{open} (i.e. Isolated) Boundary Conditions (BC), which are imposed to the system, will have on the computational evaluation of LR quantities.
We perform analytic analysis on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities which \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others which would \emph{explicitly} depend of the employed computational treatment.

@@

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed
for ground state calculations of molecules.
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will prove to be useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in Appendix~\ref{compdetails}.

@@

Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that will turn out to be very useful
in the forthcoming discussions about linear-response quantities. In particular we revisit the impact that the Isolated BC have on the solutions
of the Schr\"odinger Equation.

@@

When describing the ground state of a molecule, we suppose that the potential operator, even if non-local,
can be effectively restricted to wavefunctions that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite}
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though
perhaps in a larger region.
For a computational analysis of the problem, this property is fundamental to the choice of the numerical treatment as we know that
computational basis sets
which are tailored to express asymptotically vanishing wavefunctions are better suited for the solution of the problem.

@@

From the above expressions we can see that, for molecular systems,the value $\epsilon=0$
represents a \emph{threshold}~\footnote{The threshold value in general should correspond to asymptotic value of the potential at infinity.}
energy that separates two different classes of solutions.
Negative-energy solutions are \emph{localized}, i.e. they exhibit bound-state behavior far away from the scatterer.
These are bound states and, by means of their normalizability, they are associated to \emph{discrete}, well-defined energies.
The bound states can be of course labeled by quantum numbers which are defined by the properties of the scatterer, namely the
symmetries of the potential $\op V$.
In a computational discretization of the Schr\"odinger equation, such states can in principle be expressed with arbitrary precision,
provided that the numerical treatment offers an adequate level of completeness.
Being discrete and well-identified, the bound-state energies are a property of the molecule and can be associated to \emph{observable}
quantities. Their energies can be compared among different treatments and it is in principle possible to
provide reference values.

@@

Direct numerical evidence of this problem has been discussed in \cite{boffi2016}, where the authors analyze the effects of a finite basis set on the
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character,
and the independence of its energy with respect to the basis set.
Furthermore, it is clearly shown that the energies of the unbound orbitals are not stable with respect to parameters like the size of the computational domain.

\section{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}
It is evident that the arguments presented above have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems which are electronically stable (i.e. no anionic or metastable configurations),
such a quantity is localized, in the sense defined in the above section. Stated otherwise,
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
In this framework we consider the unperturbed Hamiltonian $\hnot$ and assumes that
the GS density $\dmnot$ is accessible and expressed in a given computational treatment.
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system.
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part $\Liouvnot$ plus a coupling term $ \coupl$. The action of these
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ shows that the response density acts as a transition
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfies the \emph{transverse} condition
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector to the empty subspace of $\hnot$
(for more details see Appendix~\ref{LiouvillianAction}).
In other terms, the condition $\dm'_\Phi = \dm'_{\Phi\,\perp}$ holds.

@@

From equation \eqref{rhoPrimeFluctuationStateDef2} we see that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is spatially localized, \textit{i.e.} its evaluation can be restricted to a bounded real-space domain.
Even when such a locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} over the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel has an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

@@

\subsection{Localized behavior for static perturbations}
The locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain.
Any local operator, \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such a property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem.
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques,
since from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{Fig1_CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the static polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line)
of a $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated with the largest
box.}
\end{figure}

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strengths $F_j$ small enough to
preserve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results show that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to those of the unperturbed KS orbitals, which is a direct proof of the bound-state behavior of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
here we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that the observable of interest $\op O$ also can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

Two distinct regimes can be identified, on the basis of the value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to those usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment, as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

@@

These considerations prove why, when a given computer code employs established GS numerical techniques,
\emph{it is much easier to converge LR quantities for $\omega$ below IP}.
Above the IP threshold, a computational setup
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able (even implicitly) to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behavior in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
for these convergence difficulties, especially for high frequencies.
The IP threshold has an evident physical interpretation: for energies in a range higher than the ionization potential,
localized and delocalized states might be coupled by the perturbation. Obtaining convergence of the observable quantities at these energies is therefore much more challenging if one employs
numerical techniques which were originally tailored for GS calculations.

As an example in support of these arguments we compare the {dynamical polarizability $\alpha(\omega)$} of a $CO$ molecule computed in three distinct real-space computational setups, which differ in the dimension of the computational domain.
Results reported in the two panels of figure \ref{co_spectrum} show the real and imaginary part of $\alpha(\omega)$, respectively. In all the cases a typical behavior emerges: the curves are almost coincident in the below threshold regime,
whereas the effects of the computational setup are clearly visible for values of  $\omega$ which exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than those presented here~\cite{baroni2008},
which are -- as already noticed -- more than sufficient below threshold.
